---
title: "Gerrymandering in Alabama"
author: "Matthew Mills"
date: "`r Sys.Date()`"
output: html_document
editor_options:
  markdown:
    wrap: sentence
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../../docs/report") })
---

# Abstract

This project investigates the racial gerrymandering in Alabama's 2021 and 2023 congressional districts using area weighted reaggregation of census blockgroup data to convex hull and minimum bounding circle geometries. Compactness scores for each district are calculated and compared.

# Study metadata

-   `Key words`: Comma-separated list of keywords (tags) for searchability. Geographers often use one or two keywords each for: theory, geographic context, and methods.
-   `Subject`: select from the [BePress Taxonomy](http://digitalcommons.bepress.com/cgi/viewcontent.cgi?article=1008&context=reference)
-   `Date created`: date when project was started
-   `Date modified`: date of most recent revision
-   `Spatial Coverage`: Specify the geographic extent of your study. This may be a place name and link to a feature in a gazetteer like GeoNames or OpenStreetMap, or a well known text (WKT) representation of a bounding box.
-   `Spatial Resolution`: Specify the spatial resolution as a scale factor, description of the level of detail of each unit of observation (including administrative level of administrative areas), and/or or distance of a raster GRID size
-   `Spatial Reference System`: Specify the geographic or projected coordinate system for the study, e.g. EPSG:4326
-   `Temporal Coverage`: Specify the temporal extent of your study---i.e. the range of time represented by the data observations.
-   `Temporal Resolution`: Specify the temporal resolution of your study---i.e. the duration of time for which each observation represents or the revisit period for repeated observations

# Study design

This is an original study based on literature on gerrymandering metrics.

It is an exploratory study to evaluate usefulness of new gerrymandering metrics based on representativness in the convex hull and minimum bounding circle compared to the congressional district.

# Materials and procedure

## Computational environment

I plan on using package ...
for ...

```{r environment-setup, include = FALSE}
# record all the packages you are using here
# this includes any calls to library(), require(),
# and double colons such as here::i_am()
packages <- c("tidyverse", "here", "sf", "tmap", "tidycensus", "htmltools", "dplyr", "knitr", "lwgeom", "markdown")

# force all conflicts to become errors
# if you load dplyr and use filter(), R has to guess whether you mean dplyr::filter() or stats::filter()
# the conflicted package forces you to be explicit about this
# disable at your own peril
# https://conflicted.r-lib.org/
require(conflicted)

# load and install required packages
# https://groundhogr.com/
if (!require(groundhog)) {
  install.packages("groundhog")
  require(groundhog)
}

# this date will be used to determine the versions of R and your packages
# it is best practice to keep R and its packages up to date
groundhog.day <- "2025-02-01"

# this replaces any library() or require() calls
groundhog.library(packages, groundhog.day)
# you may need to install a correct version of R
# you may need to respond OK in the console to permit groundhog to install packages
# you may need to restart R and rerun this code to load installed packages
# In RStudio, restart r with Session -> Restart Session

# record the R processing environment
# alternatively, use devtools::session_info() for better results
writeLines(
  capture.output(sessionInfo()),
  here("procedure", "environment", paste0("r-environment-", Sys.Date(), ".txt"))
)

# save package citations
knitr::write_bib(c(packages, "base"), file = here("software.bib"))

# set up default knitr parameters
# https://yihui.org/knitr/options/
knitr::opts_chunk$set(
  echo = FALSE, # Run code, show outputs (don't show code)
  fig.retina = 4,
  fig.width = 8,
  fig.path = paste0(here("results", "figures"), "/")
)
```

## Data and variables

We plan on using data sources precincts20, districts21, & districts23 from the districts gpkg and blockgroups2020 from the 2020 Decennial Census.

```{r}
districts_file <- here("data", "raw", "public", "alabama", "districts.gpkg")
st_layers(districts_file)
```

### Precincts 2020

```{r}
includeMarkdown(here("data", "metadata", "precincts20.md"))
```

### Districts 2023

```{r}
districts23 <- st_read(districts_file, layer="districts23")
```

Map the districts

```{r}
d23map <- districts23 %>%
  tm_shape() +
  tm_polygons(fill_alpha = 0, col = "red") +
  tm_labels(text = "DISTRICT",
          col="red",
          bgcol = "white",
          bgcol_alpha = 0.5,
          on_surface = TRUE,
          just = c("center", "center")
          )

d23map
```

### Blockgroups 2020

-   `Title`: blockgroups2020
-   `Abstract`: vector polygon geopackage layer of census tracts and demographic data
-   `Spatial Coverage`: Alabama
-   `Spatial Resolution`: Census block groups
-   `Spatial Reference System`: EPSG:4269 NAD 1983 geographic coordinate system
-   `Temporal Coverage`: 2020 census
-   `Temporal Resolution`: n/a
-   `Lineage`: taken from U.S. Census API "pl" public law summary file using `tidycensus` in R
-   `Distribution`: U.S. Census API
-   `Constraints`: Public Domain data free for use and redistribution
-   `Data Quality`: n/a
-   `Variables`: For each variable, enter the following information. If you have two or more variables per data source, you may want to present this information in table form (shown below)
    -   `Label`: variable name as used in the data or code
    -   `Alias`: intuitive natural language name
    -   `Definition`: Short description or definition of the variable. Include measurement units in description.
    -   `Type`: data type, e.g. character string, integer, real
    -   `Accuracy`: e.g. uncertainty of measurements
    -   `Domain`: Range (Maximum and Minimum) of numerical data, or codes or categories of nominal data, or reference to a standard codebook
    -   `Missing Data Value(s)`: Values used to represent missing data and frequency of missing data observations
    -   `Missing Data Frequency`: Frequency of missing data observations

| Label | Alias | Definition | Type | Accuracy | Domain | Missing Data Value(s) | Missing Data Frequency |
|:-------:|:-------:|:-------:|:-------:|:-------:|:-------:|:-------:|:-------:|
| GEOID | ... | code to uniquely identify tracts | ... | ... | ... | ... | ... |
| P4_001N | ... | Total Population, 18 years or older | ... | ... | ... | ... | ... |
| P4_006N | ... | Total: Not Hispanic or Latino, Population of one race, Black or African American alone, 18 years or older | ... | ... | ... | ... | ... |
| P5_003N | ... | Total institutionalized population in correctional facilities for adults, 18 years or older | ... | ... | ... | ... | ... |

Acquire census block group data using `tidycensus` package.
First query metadata for the `pl` public law data series.

```{r}
census_metadata_file <- here("data", "metadata", "census2020pl_vars.csv")
if(file.exists(census_metadata_file)){
  census2020pl_vars <- read.csv(census_metadata_file)
} else {
  census2020pl_vars <- load_variables(2020, "pl")
  write.csv(census2020pl_vars, here("data", "metadata", "census2020pl_vars.csv"))
}
```

In the 2023 court cases on Alabama's redistricting, it was argued that Alabama had a racial gerrymander discriminating against Black and African American voters.
Therefore, we will analyze data on the voting age population based on Black and non-black voters.
This data is found in table P3.

Query the public law data series table P3 on “race for the population 18 years and over”.

```{r}
blockgroup_file <- here("data", "raw", "public", "block_groups.gpkg")

# if the data is already downloaded, just load it
# otherwise, query from the census and save
if(file.exists(blockgroup_file)){
  blockgroups <- st_read(blockgroup_file)
} else {
  blockgroups <- get_decennial(geography = "block group",
                               sumfile = "pl",
                               table = "P3",
                               year = 2020,
                               state = "Alabama",
                               output = "wide",
                               geometry = TRUE,
                               keep_geo_vars = TRUE)
  st_write(blockgroups, blockgroup_file)
}
```

## Prior observations

Previously, racial and voting data were analyzed for 2016 and 2020 voting districts.
Compactness was assessed using the ratio of area to perimeter squared to determine relative shape to a circle.

Prior to analysis, only metadata have been observed for each data source.

## Bias and threats to validity

This study is an exploration of the modifiable aerial unit problem (MAUP).
Compactness is arbitrarily defined and fairness (presence of a racial or political gerrymander) is also subjective.
The bounding circle and convex hull methods are an attempt to apply different metrics and analyze results to form a more robust understanding of compactness and fairness in Alabama voting districts.
However, geographic features such as natural boundaries (coastlines, rivers, mountain ranges etc.) and urban settlement patterns complicate the use of universal metrics like the ones explored in the study.

## Data transformations

### Districts 2023

Reproject CRS to ESPG: 4269 and calculate Black population.

```{r}
districts23 <- districts23 %>% st_transform(4269) %>% 
  mutate(pctBlack = round(BLACK / POPULATION * 100, 1))
```

Calculate geometry area.

```{r}
districts23 <- districts23 %>%
  mutate(area = st_area(geom))
```

### Blockgroups 2020

Reproject CRS to ESPG: 4269.

```{r}
blockgroups <- st_transform(blockgroups, 4269)
```

Extract all `P3` reporting categories with people who identify as Black.

```{r}
black_vars <- census2020pl_vars %>%
  dplyr::filter(str_detect(name, "P3"),
                str_detect(label, "Black")) %>%
  select(-concept)

black_vars %>% kable()
```

Next, sum all of the categories of populations who identify as Black.
`bgarea` calculates the geographic area of each blockgroup `Total` is the total population 18 years or older `PctBlack` is the percentage of `Total` that is Black `CheckPct` sums the calculated Black population and the white population and divides by the total representing percent of the population considered in this analysis.
This should be a close underestimate (\~100) in Alabama.

```{r}
blockgroups_calc <- blockgroups %>% 
  rowwise() %>% 
  mutate(Black = sum(c_across(all_of(black_vars$name)))) %>% 
  ungroup() %>%
  mutate(bgarea = st_area(geom),
         Total = P3_001N,
         PctBlack = Black / Total * 100,
         CheckPct = (Black + P3_003N) / Total * 100
         ) %>% 
  select(GEOID, bgarea, Black, Total, PctBlack, CheckPct)
```

Save the results.

```{r}
st_write(blockgroups_calc, 
         here("data", "derived", "public", "blockgroups_calc.gpkg"),
         append=FALSE)
```

```{r}
tmap_mode("plot")
```

And plot results to visualize.

```{r}
blkgrp_black_map <- tm_shape(blockgroups_calc) + 
  tm_polygons(
    fill = "PctBlack",
    col_alpha = 0.2,
    lwd = 0.1,
    col = "grey90"
  )

blkgrp_black_map
```

Now, we visualize the 2023 districts overlaid on blockgroups by pctBlack

```{r}
tmap_mode("view")
```

```{r}
blkgrp_black_map +
  d23map
```

## Analysis


Use area weighted reaggregation (AWR) to determine the voting age Black and total populations in each district using from st_intersection().
Next, AWR will be repeated on the convex hull and minimum bounding circle geometries.
Compactness scores will be calculated for each analysis geometry.

Research is exploratory so there are no statistical tests or weighting criteria.

### 2023 districts

Estimate voting populations by race for each 2021 district

```{r}
d23_estimate <- st_intersection(blockgroups_calc, districts23) %>%
  mutate(awRatio = st_area(geom)/bgarea,
         awTotal = awRatio * Total,
         awBlack = awRatio * Black) %>%
  st_drop_geometry() %>%
  group_by(DISTRICT) %>%
  summarize(bgBlack = sum(awBlack), bgTotal = sum(awTotal))

districts23_join_bg <- districts23 %>% 
  left_join(d23_estimate, by = "DISTRICT") %>% 
  mutate(pctBlackbg = round(bgBlack / bgTotal * 100, 1))
```

report results

```{r}
districts23_join_bg %>% st_drop_geometry() %>% kable()
```
Estimate voting populations by race using convex hull and join to blockgroup estimates

```{r}
d23_estimate <- st_intersection(blockgroups_calc, st_convex_hull(districts23)) %>%
  mutate(awRatio = st_area(geom)/bgarea,
         awTotal = awRatio * Total,
         awBlack = awRatio * Black) %>%
  st_drop_geometry() %>%
  group_by(DISTRICT) %>%
  summarize(bgBlack = sum(awBlack), bgTotal = sum(awTotal))

districts23_join_ch <- districts23_join_bg %>% 
  left_join(d23_estimate, by = "DISTRICT") %>% 
  mutate(pctBlackch = round(bgBlack.y / bgTotal.y * 100, 1),
         diffPct = as.numeric(pctBlackbg) - as.numeric(pctBlackch),
         absdiffPct = abs(diffPct))
```

Estimate voting populations by race using minimum bounding circle and join to blockgroup estimates

calculate compactness scores for each geometry
```{r}
districts23_results <- districts23_join_ch %>%
  mutate(
    darea = st_area(geom),
    dperim = st_length(st_cast(geom, "MULTILINESTRING")),
    compact_shp = round(as.numeric(4 * pi * darea / dperim^2),2),
    compact_ch = round(as.numeric(darea / st_area(st_convex_hull(geom))),2),
    compact_mbc = round(as.numeric(darea / st_area(st_minimum_bounding_circle(geom))),2)
  )
```

# Results

### Correlation matrix and small plots by gerrymandering indicators

```{r}
districts23_results_cor <- districts23_results %>% 
  st_drop_geometry() %>%
  select(pctBlackbg,
         diffPct,
         absdiffPct,
         compact_shp,
         compact_ch,
         compact_mbc)

districts23_results_cor %>% pairs()
```
```{r}
districts23_results_cor %>% cor() %>% kable()
```

### Plot difference in representation (between district and convex hull) against compactness scores

```{r}
districts23_results_plot <- districts23_results %>% 
  st_drop_geometry() %>%
  select(DISTRICT, pctBlack, absdiffPct, compact_shp, compact_ch, compact_mbc) %>% 
  pivot_longer(cols = starts_with("compact"))

districts23_results_plot %>% ggplot() +
  aes(x = absdiffPct, y = value) +
  geom_smooth(method="lm", col = "grey30") +
  geom_label(aes(label = DISTRICT, fill = pctBlack)) +
  scale_fill_distiller(type = "div", palette = "PRGn") +

  facet_wrap(~name)
```

There is a negative correlation between compactness score and percent absolute difference for both the convex hull and shapefile geometry calculations. The deviation from the regression is also relatively small, indicating that there is more accuracy to this trend. The bounding circle results, however, were less consistent. They showed a slight positive increase in absolute difference. Compactness score for district 5 decreased while scores for districts 4, 6, and 7 increased. Looking at district 5, it sits along a flat, northern boundary. Because it does not extend north, its score decreases despite the relative lack of obvious gerrymandering. For districts 4, 6, and 7, we can see how obvious deviations in their geometries are hidden in the boundary circle method as they contain deviations in different directions that make their overall shape more circular.

```{r}
tm_shape(districts23_results) +
  tm_polygons(fill = "pctBlackbg") +
  tm_text("DISTRICT")
```

# Discussion

Negative correlation between compactness score and absolute difference indicates a gerrymander. Districts that were the least compact were also the least representative. In a world without gerrymandering, differences in compactness would correspond with natural boundaries and thus not be associated with less representation. Bounding circle results yielded the least compelling results as they were unable to detect and represent clear intentional gerrymandering in district 7 and possibly districts 4 and 6 as well. Ultimately, we see that racial gerrymandering remains in 2023 despite redistricting.

# Integrity Statement

The authors of this preregistration state that they completed this preregistration to the best of their knowledge and that no other preregistration exists pertaining to the same hypotheses and research.

# Acknowledgements

This report is based upon the template for Reproducible and Replicable Research in Human-Environment and Geographical Sciences, [DOI:[10.17605/OSF.IO/W29MQ](DOI:%5B10.17605/OSF.IO/W29MQ){.uri}](<https://doi.org/10.17605/OSF.IO/W29MQ>)

# References
